\chapter{Ordinary Differential Equations}

Dynamical systems, of natural, technological or mixed structure are very often described in terms of \odes.
A differential equation is an equation involving derivatives of an unknown function that depends upon one or more independent variables. \cite{tang_2007}. If the unknown function depends on only one independent variable, then the equation is called an ordinary differential equation.

An \ode can be either linear or non-linear. The latter contains a non-linear expression of the unknown function, in a linear \ode all expressions of the unknown function (including its derivatives) are linear. Though non-linear \odes play a great role in the description of dynamical processes, in this treatise we concentrate on linear dynamical systems and therefore restrict our attention on the linear \ode.

An \ode can be represented in the following standard form:
\begin{equation}\label{eq: linear ode general}
	a_n \frac{d^ny}{dt^n} + a_{n-1} \frac{d^{n-1}y}{dt^{n-1}} + ... + a_1 \frac{dy}{dt} + a_0 y = f(t),
\end{equation}
where $y = y(t)$ is the unknown function, $a_i = a_i(t)$ the coefficients of the \ode, $f(t)$ a known external function. The value $n$ which is supposed to be different from zero is the order of the \ode. To solve the equation (\ref{eq: linear ode general}) one also needs to define the initial conditions, i.e. the values $y_0, y'_0, ..., y_0^{(n-2)}, y_0^{(n-1)}$ of the function and its derivatives up to the order $n-1$ at the moment of time $t=0.$

The coefficients $a_i$ can be either functions of time or constant. In the latter special case we have to do with a linear \ode with constant coefficients. Analytical solutions of linear \odes cannot be obtained in the general case, whereas in the case of \odes with constant coefficients it can be always done in compact form. Fortunately, many interesting and practical dynamical phenomena in nature and technology can be described through linear \ode with constant coefficients. In this treatise we will primarily concentrate on those.

\section{Homogeneous equation}
If in equation (\ref{eq: linear ode general}) $f(t) = 0$, this is, there is no external force acting,  the \de is called homogeneous. Its general solution can be written down as a sum of basic functions $\varphi_i(t)$:
\begin{equation}\label{eq: basic functions}
	y(t) = \sum \limits_{i=1}^n C_i \varphi_i(t),
\end{equation}
where $C_i$ are constant coefficients that need to be uniquely defined using the initial conditions. The basic functions $\varphi_i(t)$ depend on the roots of the algebraic characteristic equation
\begin{equation}\label{eq:characteristic equation}
	a_n \lambda^n + a_{n-1} \lambda^{n-1} + ... + a_1 \lambda + a_0 = 0.
\end{equation}
As is known from algebra, the equation (\ref{eq:characteristic equation}) has exactly $n$ roots. Some of them can be realm some complex (in which case there will always be pairs of conjugated roots); the equation (\ref{eq:characteristic equation}) can also the same root occurring more than once.
	\paragraph{Examples\\}
	Characteristic equation $\lambda^2 + \lambda -2 = 0$ has two different real roots, $\lambda_1 = 1$ and $\lambda_2 = -2$. 
	
	Characteristic equation $\lambda^2 -2 \lambda + 5 = 0$ has two conjugated roots, $\lambda_1 = 1 + 2 \iu$ and $\lambda_2 = 1 - 2 \iu$, where $\iu = \sqrt{-1}$.
	
	Characteristic equation $\lambda^4 + 3 \lambda^3 - 3 \lambda^2 - 11 \lambda - 6 = (\lambda + 1)^2 (\lambda + 3) (\lambda - 2) = 0$
 	has two equal roots, $\lambda_1 = \lambda_2 = -1$ and two different other roots $\lambda_3 = -3$ and $\lambda_4 = 2$.
 	
 	Characteristic function $\varphi_i(t)$ in (\ref{eq: basic functions}) depends oh the character of the root $\lambda_i$:
 	\begin{itemize}
 		\item If $\lambda_i$ is a simple root of (\ref{eq: basic functions}), $\varphi_i(t) = \E^{\lambda_i t}$
 		\item If $\lambda_i$ and $\lambda_{i + 1}$ are conjugated complex roots of  (\ref{eq: basic functions}), $\varphi_i(t) = \E^{\lambda_i t}$ and $\varphi_{i}+1(t) = \E^{\lambda_{i+1} t}$ are basic functions 
 		\item If $ \lambda $ is a repeated root, witn $m$ occurrences (e.g. $\lambda_1 = \lambda, \lambda_2 = \lambda, ... \lambda_m = \lambda$), the basic functions are $\varphi_1(t) = \E^{\lambda t}, \varphi_2(t) = t \E^{\lambda t}, ..., \varphi_m(t) = t^{m-1} \E^{\lambda t} $ \cite{tang_2007}.
 	\end{itemize}
 	
In many practical cases, however, repeated roots of the characteristic equation are rather seldom, so one can suppose that all roots are different, either real or complex, in which case there are always two conjugated ones.

	\paragraph{Examples. Simple linear \odes}
	\subparagraph{Case 1. Two different real roots of the \che} Our \de is
	\begin{equation}\label{eq: simple linear ode example 1}
		\frac{d^2y}{dt^2} + \frac{dy}{dt} - 2 y = 0,	
	\end{equation}
	initial conditions: $y(0) = 0, ~y'(0) = 1$.\\
	\textbf{Solution}. The \che of this \de is $\lambda^2 + \lambda - 2 = 0$, which, as we have seen above has the roots $\lambda_1 = 1$ and $\lambda_2 = -2$. Therefore the general solution of this \de will be
	\[ 
		y(t) = C_1 \E^t + C_2 \E^{-2t},
	\]
	where $C_1, C_2$ are constants that need to be calculated using the initial conditions. We have $y(t) = C_1 + C_2$, $y'(t) = C_1 \E^t - 2 C_2 \E^{-2t}$, $y'(0) = C_1 - 2 C_2$. \\
	From the initial conditions we have $C_1 + C_2 = 0,~C_1 - 2 C_2 = 1. $ Solving these equations, we have $C_1 = 1,~C_2 = -1.$ The solution in final form is therefore $y(t) = \E^t - \E^{-2t}.$
	
	\subparagraph{Case 2. Two complex conjugated roots of the \che}
	Let the \de now be 
	\begin{equation}\label{eq:}
		\frac{d^2y}{dt^2} - 2 \frac{dy}{dt} + 5 y = 0,
	\end{equation}
	
	initial conditions: $y(0) = 0, ~y'(0) = 1$.\\
	\textbf{Solution}. The \che of this \de is $\lambda^2 -2 \lambda + 5 = 0$, which, as we have seen above, has the roots $\lambda_1 = 1 + 2 \iu$ and $\lambda_2 = 1 - 2 \iu$. \\
	Therefore the general solution of this \de will be
	\[ 
		y(t) = C_1 \E^{(1 + 2 \iu)t} + C_2 \E^{(1 - 2 \iu)t}.
	\]
	
	From the initial conditions, we have $C_1 + C_2 = 0, ~ (1 + 2 \iu) C_1 + (1 - 2 \iu) C_2 = 1$ whence $\ds C_1 = \frac{1}{\lambda_1 - \lambda_2} = \frac{1}{4 \iu},~  C_2 = - \frac{1}{4 \iu}$ and
	\[ 
	y(t) = \frac{1}{4 \iu} \left[ \E1{(1 + 2 \iu) t } - \E1{(1 - 2 \iu) t } \right] = \frac{1}{4 \iu} \E^t \left( \E^{2\iu t}  -  \E^{-2\iu t} \right).
	\]
	
	Since\cite{Abramowitz1964}  $\ds \frac{1}{2 \iu} \left( \E^{2\iu t}  -  \E^{-2\iu t} \right) = \sin 2t$ , we obtain that $y(t) = \frac{1}{2} \E^t \sin 2t$.
	
	\section{Non-homogeneous equation}\label{sec:Non-homogeneous equation}
	In the case when $f(t) \neq 0$ in the right part of (\ref{eq: linear ode general}) there is no receipt to obtain the solution straightforwardly. The general idea is to find a particular solution of (\ref{eq: linear ode general}) and to combine it with the solution of the homogeneous equation \cite{tang_2007} in the form of a linear combination with undefined coefficients, The latter can then be computed out of the initial conditions. 
	
	For instance, if the right-part function is a polynomial, a partial solution of \eqref{eq: simple linear ode example 1} can be also looked for in the form of a polynomial of $t$ \cite{tang_2007}. For example, for the equation \eqref{eq: simple linear ode example 1} a partial solution can be looked for in the form of $\tilde{y}(t) = a t^2 + bt + c.$ Since $\tilde{y}'(t) = 2 a t + b$ and $\tilde{y}''(t) = 2a$, substituting into \eqref{eq: simple linear ode example 1} we obtain that the partial solution is 
	\begin{equation}\label{eq:partial solution example 1}
		\tilde{y}(t) = -\frac{k}{2} t - \frac{1}{4}.
	\end{equation}
	The general solution of \eqref{eq: simple linear ode example 1} needs to be looked for as 
	\[ 
		y(t) = C_1 \E^t + C_2 \E^{-2t} + \tilde{y}(t).
    \]
    
    From the initial conditions $y(0) = 0, ~y'(0) = 1$ we have that 
    \[ 
      \begin{aligned}
    	C_1 + C_2 - \frac{1}{4} k  &= 0 \\
    	C_1 - 2C_2 - \frac{1}{12} k &= 1.
      \end{aligned}
    \]
	
	Solving the above system of two linear equations with respect to $C_1, C_2$ we obtain that
	\[ 
		C_1 = \frac{k + 1}{3}, ~C_2 = -\frac{1}{3}\left( 1 + \frac{k}{4} \right)
	\]
	and
	\[ 
		y(t) = \frac{k + 1}{3} \E^t -\frac{k + 4}{12} \E^{-2t} - \frac{k}{4} (2t + 1).
	\]
	
	No general recipe to obtain a particular solution for an arbitrary right part function $f(t)$ however exists. Textbooks only provide methods for a few classes of right-part functions, like the polynomial case in the example above. 
	
	A general method of solving linear \odes with constant coefficients and arbitrary external forcing $f(t)$ is given by the \lt which will be the subject of the next chapter.
	